---
output: 
  rmarkdown::html_document:
    css: custom.css
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r test_setup, echo = FALSE, message=FALSE}
options(knitr.table.format = "html")

    # df_print: paged
# <script async defer src="https://hypothes.is/embed.js"></script>
library(dplyr)
library(ecosscraper)
library(fuzzyjoin)
library(ggplot2)
library(kableExtra)
library(knitr)
library(leaflet)
library(parallel)
library(rio)
library(rmapshaper)
library(stringr)
library(tokenizers)
library(us.geonames)

# read in some data
cnt_occ <- readRDS("data/ESA_county_occurrences.rds")
fiveyr_table <- readRDS("data/fiveyr_table_2016-12-17.rds")

# TECP_table <- get_TECP_table()
tne <- TECP_table %>% filter_listed() %>% filter(U_S__or_ForeignListed != "Foreign")

get_text <- function(file) {
    text <- readLines(con = file) %>%
      paste(collapse = " ") %>%
      str_replace_all("[ ]{2,}", " ")
    return(text)
}
```

# Extinction {#extinct}

Extinction is final: absent leaps in genomic engineering, once a species is
gone it is gone for good. The extinction crisis of the modern era was a key
motvation for the ESA (CITES): the goals of the law include preventing 
extinction and recovering species so they are no longer threatened with
extinction. One of the ways the success of the ESA is measured is by comparing
the number of listed species to the number that are extinct. As of this writing,
ten species have been removed from the list of threatened and endangered
species, and the list includes 1,652 species. Those numbers suggest 
`r 100 - (100 * round(10 / 1652, 3))`% of species have been saved by the ESA,
which by any means is a superb accomplishment.

There are a few problems with using the number of species "delisted due to 
extinction" as the metric of success. First, it ignores the key goal of species
recovery and uses the lower bar of mere existence as success. But would anyone
really argue that plants and animals known only to exist in captivity count as
successes?

Second, we know that the ten delisted species are simply the ones formally
accepted as extinct. But proving extinction is tough; we have too many examples
where a species hadn't been seen for many, many years--multiple decades and even
over a century--only to be rediscovered (CITES). For such reasons, declarations 
of extinction are a rare occurrence. At the same time, we often have a  darn 
good idea that well-documented species are really-and-truly extinct. We hold off
on the declarations of extinction and the removal of protections because of the
small but real chance that one or a few individuals still exist. 

My goal here is to find the species that experts - biologists with the U.S. Fish 
and Wildlife Service (FWS) and National Marine Fisheries Service (NMFS; 
collectively, the Services) - have good reason to believe are or may be extinct.
_Declaring_ extinction may not be warranted for many or most of these, but 
these professionals have indicated that extinction is likely to some greater or
lesser degree. Getting a better handle on the number of species thought to be
extinct rather than relying on the number delisted due to extinction is an 
essential component to understanding the effectiveness of the ESA.

## Identifying possible/probably extinctions

The first question we address is how many species are extinct. Pinning
down that number is tricky because extinction is fundamentally hard because 
proving absence is so difficult (CITES). Rather than a declaration of extinct,
we set the bar at a level of experts' statements that they think the species
is or may be extinct. From among the ~1,400 five-year reviews, totalling >30,000 
pages, we need to find a relatively small number of phrases that indicate 
extinction. We used a set of seven [regular expression](https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf) patterns as a coarse filter:

<div style="font-size:smaller">
1. `"(possibly|may be|have become) extinct"`
2. `"(certainly being|probably|is|likely|probably being) extinct"`
3. `"no (populations|individuals)( are)* known"`
4. `"no( known| living)* individuals"`
5. `"extinct in( the)* wild"`
6. `"last (seen|observed|found) in [0-9oOiI]+"`
7. `"functionally extinct"`
</div>
    
In these patterns, "|" means "or" for the set of words inside parentheses. For
example, the first pattern would match "possibly extinct," "may be extinct," or
"have become extinct." The "*" means the preceding letter or word matches zero
or more times, whereas "+" means the preceding word/letter(s) must match one or
more times. We checked all seven patterns against all 1,385 five-year reviews
we collected from [ECOS](https://ecos.fws.gov) and [NMFS's recovery site](http://www.nmfs.noaa.gov/pr/recovery/plans.htm).

```{r terms_1, echo = FALSE}
src <- "~/Downloads/five_year_review"
fils <- list.files(src, full.names = TRUE, recursive = TRUE)

terms <- c(
  "(possibly|may be|have become) extinct",
  "(certainly being|probably|is|likely|probably being) extinct",
  "no (populations|individuals)( are)* known",
  "no( known| living)* individuals",
  "extinct in( the)* wild",
  "last (seen|observed|found) in [0-9oOiI]+",
  "functionally extinct"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- get_text(f) %>%
    tokenize_sentences(lowercase = TRUE, simplify = TRUE)
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(terms)) {
    mat <- grep(text, pattern = terms[i], fixed = FALSE, value = TRUE)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

# t1 <- parallel::mclapply(fils, 
#                          proc_file, 
#                          mc.cores = 3, 
#                          mc.preschedule = FALSE)
# ext_pat <- bind_rows(t1) %>% filter(!is.na(mats))
# fname <- paste0("data/extinct_patt_matches-", Sys.Date(), ".rds")
# saveRDS(ext_pat, fname)
fname <- paste0("data/extinct_patt_matches-2017-09-12.rds")
ext_pat <- readRDS(fname)
```

This search returned `r dim(ext_pat)[1]` matches to the patterns, with pattern
six ("last seen in...") most common:

<div class="narrow-tab">
```{r echo = FALSE} 
table(ext_pat$patt) %>%
  as_data_frame() %>%
  rename(pattern = Var1) %>%
  kable() %>%
  kable_styling(
    bootstrap_options = "hover", 
    full_width = T
  )
```
</div>

A few examples of sentences in which pattern six was found include:

<div class="narrow-tab">
```{r patt_6, echo = FALSE} 
t3 <- filter(ext_pat, patt == 6) %>% 
  select(mats) %>% 
  data.frame() %>%
  head(10)
kable(paste("-", t3$mats))
```
</div>

And examples of pattern two (higher likelihood of extinction) matches include:

<div class="narrow-tab">
```{r patt_2, echo = FALSE} 
t4 <- filter(ext_pat, patt == 2) %>% 
  select(mats) %>% 
  data.frame() %>%
  head(10)
kable(paste("-", t4$mats))
```
</div>

Perusing the pattern matches, we find (as expected) both true positives (i.e.,
the meaning of the matched sentence comports with the idea we have in mind) and
false positives (e.g., local extinctions or species previously believed extinct
but since re-discovered). We also recognize that the current table doesn't
identify the (one or more) species in each document. After joining the pattern
matches with a table that links five-year review documents and species
(scientific names), we manually checked each of the `r unique(ext_pat$doc) %>% length()` five-year reviews with matches to 
# filter out the false-positives. 

```{r load_manual, echo = FALSE}
# fiveyr_table <- readRDS("data/fiveyr_table_2016-12-17.rds")
# w_spp <- left_join(ext_pat, fiveyr_table, by = "doc")
# rio::export(w_spp, "data/ext_patt_w_spp.xlsx")
#
###### MANUAL FILTER #######
#
ext_5yr <- readRDS("data/extinct_manual_check_2017-09-05.rds")

extinct <- filter(ext_5yr, true_positive == "T")
```

From the manually filtered data we find `r length(unique(extinct$Species))`
species that are believed extinct to some greater or lesser degree, plus the 
snail genus _Achatinella_. There were 41 species listed in the snail genus and a 
number - somewhere between 15 and 30 species, though it is difficult to tell 
exactly how many - are thought to be extinct.

<div class="cols-2">
```{r echo=FALSE}
spp <- sort(unique(extinct$Species[!is.na(extinct$Species)]))
half <- ceiling(length(spp) / 2)
first <- spp[1:half]
secnd <- spp[(half+1):length(spp)]
kable(first)
kable(secnd)
```
</div>

<div class="main-result">
Together with the ten species that have been [delisted due to extinction](https://ecos.fws.gov/ecp0/reports/delisting-report), **we estimate 70-85 (`r 100*round(70/dim(tne)[1], 2)` -  `r 100*round(85/dim(tne)[1], 2)`%) current or formerly ESA-listed species are believed, to some greater or lesser extent, to be extinct.**
</div>

We address the geography of extinctions below, but note here that most of these
species are from Hawaii and many of those are plants (Box 1).

<div class="aside-box">
### Box 1. Complementary data from PEP

The [Plant Extinction Prevention (PEP) Program](http://www.pepphi.org/) works
exhaustively to monitor and conserve Hawaiian plants. They also provide more up-
to-date data on the status of many imperiled plants than FWS documents, and 
should probably be considered authoritative for PEP species.

We downloaded the 2017-03-16 list from PEP's website then used Adobe's online
extraction service to render the tables as a spreadsheet for processing. To
begin, we have a high-level overview of how PEP classifies the covered species:

<div class="narrow-tab">
```{r echo=FALSE}
PEP <- readRDS("data/HI_PEP_2017-03-16.rds")
extrp_wild <- filter(PEP, island_designation == "PEP-EXTIRPATED" |
                       island_designation == "PEP-EXTIRPATED?")
table(PEP$island_designation) %>%
  as_data_frame() %>%
  rename(PEP_class = Var1) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

Appropriately, PEP uses "?" to denote species for which extirpation and 
extinction are unknown. The refinement of extinct in the wild vs. extinct from
the planet is also useful to consider. Last, in addition to the extinctions, the 
PEP data highlight the problem of extirpations: `r length(unique(extrp_wild$species))` 
PEP species have been (or probably have been) extirpated from one or more 
of the islands. First, the number of extirpations by island:

<div class="narrow-tab">
```{r echo=FALSE}
spXis_tab <- table(extrp_wild$species, extrp_wild$distribution)
colnames(spXis_tab) <- str_replace_all(colnames(spXis_tab), "`", "")
colSums(spXis_tab) %>% 
  sort(decreasing = TRUE) %>% 
  head(10) %>% 
  as_data_frame() %>%
  rename(extirpations = value) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

And second, by most species:

<div class="narrow-tab">
```{r echo = FALSE}
rowSums(spXis_tab) %>%
  sort(decreasing = TRUE) %>% 
  head(10) %>%
  as_data_frame() %>%
  rename(extirpations = value) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

```{r echo=FALSE}
hi_cnt <- filter(cnt_occ, STATE == "Hawaii")
pep_miss <- setdiff(unique(hi_cnt$SCI), PEP$species)
```

The PEP data adds depth to our understanding of extinctions and extirpations
among Hawaiian plants, but PEP doesn't cover all ESA-listed plants.  By 
combining PEP table with county occurrence data, we find `r length(pep_miss)` 
species are not covered by PEP; more are missing than are covered by PEP. For
these species in particular it is important for FWS to track species status. 
One way that is done is with ESA-mandated five-year reviews of species status.

```{r echo=FALSE, message=FALSE, warning=FALSE}
pep_miss_5y <- filter(fiveyr_table, fiveyr_table$Species %in% pep_miss)
pep_miss_5y$age <- as.numeric(Sys.Date() - as.Date(pep_miss_5y$Date))
date_tab <- as.Date(pep_miss_5y$Date) %>%
  lubridate::year() %>%
  table() %>%
  as_data_frame() %>%
  rename(year = '.')
p <- ggplot(date_tab, aes(x = year, y = n)) +
  geom_histogram(stat = "identity") +
  labs(x = "", y = "count") +
  ggthemes::theme_hc()
plotly::ggplotly(p)
```

There are `r length(pep_miss_5y$age)` non-PEP species with 5-year 
reviews available, and of those reviews, `r sum(pep_miss_5y$age > (365*5))` are 
out-of-date (i.e., more than five years old).
</div>

## Geography of extinctions

<div class="narrow-tab">
```{r state_level, echo = FALSE}
w_co <- left_join(extinct, cnt_occ, by = c("Species" = "SCI"))
spp_w_st <- distinct(w_co, Species, STATE)

dist <- filter(ext_5yr, true_positive == "T") %>% 
  distinct(Species, .keep_all = TRUE) %>%
  filter(!is.na(Date))

ext_geo <- left_join(dist, cnt_occ, by = c("Species" = "SCI"))
ext_geo_state <- distinct(ext_geo, Species, STATE, .keep_all = TRUE)
table(ext_geo_state$STATE) %>%
  sort(decreasing = TRUE) %>%
  as_data_frame() %>%
  rename(state = Var1) %>%
  head(10) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

----

<div class="narrow-tab">
```{r county_level, echo = FALSE}
ext_geo$co_st <- paste0(ext_geo$NAME, ", ", ext_geo$STATE)
ext_geo_st_co <- distinct(ext_geo, Species, co_st, .keep_all = TRUE)
ext_state_cnt <- table(ext_geo_state$STATE) %>% 
  sort(decreasing = TRUE) %>%
  as_data_frame()
names(ext_state_cnt) <- c("state", "n_extinct")
table(ext_geo_st_co$co_st) %>% 
  sort(decreasing = TRUE) %>% 
  as_data_frame() %>%
  rename(state_county = Var1) %>%
  head(15) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

## Common threats and habitats

