---
output: 
  rmarkdown::html_document:
    css: custom.css
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r test_setup, echo = FALSE, message=FALSE}
options(knitr.table.format = "html")
library(plyr)

library(bsplus)
library(dplyr)
library(ecosscraper)
library(fuzzyjoin)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(knitr)
library(leaflet)
library(parallel)
library(rredlist)
library(rio)
library(rmapshaper)
library(RPostgreSQL)
library(stringr)
library(tokenizers)
library(us.geonames)

use_bs_tooltip()
use_bs_popover()

# read in some data
cnt_occ <- readRDS("data/ESA_county_occurrences.rds")
fiveyr_table <- readRDS("data/fiveyr_table_2016-12-17.rds")

# TECP_table <- get_TECP_table()
# tne <- TECP_table %>% filter_listed() %>% filter(U_S__or_ForeignListed != "Foreign")

con <- dbConnect(
  dbDriver("PostgreSQL"),
  dbname = "esc-dev",
  user = "postgres",
  password = "SillySilly1!",
  host = "localhost"
)

tne <- dbSendQuery(con, "select * from tecp_table") %>%
  dbFetch() %>%
  filter_listed() %>%
  filter_domestic()

get_text <- function(file) {
    text <- readLines(con = file) %>%
      paste(collapse = " ") %>%
      str_replace_all("[ ]{2,}", " ")
    return(text)
}

simple_tab <- function(v, vname = "var", useNA = "always", n = 10) {
  table(v, useNA = useNA) %>% 
    sort(decreasing = TRUE) %>%
    as_data_frame() %>%
    plyr::rename(replace = c("v" = vname)) %>% 
    head(n) %>%
    kable() %>%
    kable_styling(bootstrap_options = "hover")
}
```

# Extinction and the ESA {#extinct}

<div class="abstract">
**_Abstract_**

The success of the Endangered Species Act (ESA) is often measured by comparing the number of species declared extinct to the number of listed species, which indicates success on the order of 99%. Here we check that measure against the discussions of extinction by the Fish and Wildlife Service and the National Marine Fisheries Service by mining five-year status reviews and other relevant data. The review indicates 89 (~5%) domestic ESA-listed species are known or are likely to be extinct, plus at least six and likely more that are extinct in the wild. We are unable to provide a rigorous analysis of causes of extinction, but identify three broad categories of common causes: invasive species, natural accidents, and overexploitation / habitat destruction. We discuss how these results inform messages about the successes of the ESA and the law's limitations, and close with a hopeful but cautionary tale about extinction. 
</div>

The extinction crisis of the modern era was a key motivation for the ESA: the goals of the law include preventing extinction and recovering species so they are no longer threatened with extinction. One of the ways ESA success has been measured is by comparing the number of listed species to the number that have been declared extinct and delisted. As of this writing, ten species have been removed from the list of threatened and endangered species because of extinction, and the list includes 1,660 species for approximately `r 100 - (100 * round(10 / 1660, 2))`% success.

But there are at least two problems with using the number of species "delisted due to extinction" as the metric of success. First, it sets a lower bar of mere existence as success rather than keeping the higher goal of recovery. But we don't want to argue that plants and animals known only to exist in captivity, or in such low numbers that their [ecological role](https://smile.amazon.com/Ecological-Niches-Contemporary-Interspecific-Interactions/dp/0226101800/ref=sr_1_1) is lost, count as successes. Success is wildlife out on the landscape as part of the ecological community.

Second, we know that the ten delisted species are just the ones that have been formally declared extinct. But proving extinction is tough: it's easy to confirm something is present, but confirming absence (of typically hard-to-find species) is fraught with problems. Did we look at the right place, at the right time, and under the right conditions and just miss it? Or is the species truly gone? The caution of declaring extinction is well-deserved: we have too many examples where a [species hadn't been seen for decades or even over three centuries](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0022531), only to be rediscovered. And getting either extinction or rediscovery wrong can have [negative consequences for conservation](https://www.researchgate.net/publication/26879484_Caution_with_claims_that_a_species_has_been_rediscovered). For such reasons, declarations of extinction are a rare occurrence. At the same time, we often have a very good idea that well-documented species are extinct, but hold off on the declarations of extinction and the removal of protections because of the small chance that one or a few individuals still exist. 

We have two goals in this analysis. First, we identify ESA-listed species that experts--primarily biologists with the U.S. Fish and Wildlife Service (FWS) and National Marine Fisheries Service (NMFS; collectively, the Services)--have good reason to believe are extinct. We use a [combination of computational and manual curation](#methods) and data from other highly regarded sources, including IUCN and the [Plant Extinction Prevention Program](http://pepphi.org), to create this list. _Declaring_ extinction may not be warranted for most of these species, but the professionals have indicated that extinction is likely to some greater or lesser degree. Second, we summarize the likely causes of these species' extinction. This is not a rigorous analysis because of various uncertainties, but it provides an important (and somewhat surprising) backdrop for interpreting the extinction data. We discuss the implications of our results with respect to how conservationists discuss the effectiveness of the ESA.

## Extinctions

```{r terms_1, echo = FALSE}
src <- "~/Downloads/five_year_review"
fils <- list.files(src, full.names = TRUE, recursive = TRUE)

terms <- c(
  "(possibly|may be|have become) extinct",
  "(certainly being|probably|is|likely|probably being) extinct",
  "no (populations|individuals)( are)* known",
  "no( known| living)* individuals",
  "extinct in( the)* wild",
  "last (seen|observed|found) in [0-9oOiI]+",
  "functionally extinct"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- get_text(f) %>%
    tokenize_sentences(lowercase = TRUE, simplify = TRUE)
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(terms)) {
    mat <- grep(text, pattern = terms[i], fixed = FALSE, value = TRUE)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

# t1 <- parallel::mclapply(fils, 
#                          proc_file, 
#                          mc.cores = 3, 
#                          mc.preschedule = FALSE)
# ext_pat <- bind_rows(t1) %>% filter(!is.na(mats))
# fname <- paste0("data/extinct_patt_matches-", Sys.Date(), ".rds")
# saveRDS(ext_pat, fname)
fname <- paste0("data/extinct_patt_matches-2017-09-12.rds")
ext_pat <- readRDS(fname)
```

```{r load_manual, echo = FALSE}
# fiveyr_table <- readRDS("data/fiveyr_table_2016-12-17.rds")
# w_spp <- left_join(ext_pat, fiveyr_table, by = "doc")
# rio::export(w_spp, "data/ext_patt_w_spp.xlsx")
#
###### MANUAL FILTER #######
#
ext_5yr <- readRDS("data/extinct_manual_check_2017-09-05.rds")

extinct <- filter(ext_5yr, true_positive == "T") %>%
             filter(is.na(comment) | 
                      comment == "HI birds" |
                      comment == "uncertain status") %>%
             filter(Species != "Phyllostegia mollis") %>%
             filter(Species != "Melicope quadrangularis") %>%
             filter(Species != "Cyanea superba") 
spp_ls <- unique(extinct$Species[!is.na(extinct$Species)])
spp_ls <- c(spp_ls, "Campephilus principalis", "Vermivora bachmanii")

ext_wild <- filter(ext_5yr, comment == "extinct in wild") %>%
  filter(Species != "Cyanea superba") %>%
  distinct(Species)
```

Using [full-text searches plus manual filtering](#methods), we first identified `r length(spp_ls) - 1` species that are believed extinct to some greater or lesser degree, plus the [snail genus _Achatinella_](#achatinella).

<div class="cols-2">
```{r echo=FALSE}
spp <- sort(spp_ls)
half <- ceiling(length(spp) / 2)
first <- spp[1:half]
secnd <- spp[(half+1):length(spp)]
kable(first)
kable(secnd)
```
</div>

The list includes some well-known species, like the [eastern cougar](https://www.fws.gov/northeast/ECougar/) and the [Eskimo Curlew](https://neotropical.birds.cornell.edu/Species-Account/nb/species/eskcur/overview). But other species are much less well-known, like _Cyrtandra crenata_, a Hawaiian plant that was last seen in 1947. The list also includes the only listed genus, _Achatinella_, the Oahu tree snails.

### _Achatinella_ snails {#achatinella}

```{r echo=FALSE}
# acha <- readxl::read_excel("data-raw/Achatinella.xlsx")
# saveRDS(acha, file = "data/Achatinella.rds")
acha <- readRDS("data/Achatinella.rds")
acha$species <- paste("Achatinella", acha$species)
ext_ach <- filter(acha, believed_extinct == "T")
```

Some `r length(unique(acha$species))` species or subspecies of _Achatinella_ snails have been described, all native to the tropical forests of Oahu. Unfortunately, they have suffered more likely extinctions than any other genus of ESA-listed species. Based on the status discussed in the last five-year reviews for the genus (2011) plus information gleaned from other sources (e.g., IUCN and NatureServe), `r length(ext_ach$species)` _Achatinella_ are likely extinct:

<div class="narrow-tab">
```{r echo=F}
select(ext_ach, c(1,2,9)) %>%
  kable(align = c("l", "c", "c")) %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

### Plant Extinction Prevention (PEP) Program

The [Plant Extinction Prevention (PEP) Program](http://www.pepphi.org/) works exhaustively to monitor and conserve Hawaiian plants. They also provide more up-to-date data on the status of many imperiled plants than FWS documents, and should be considered authoritative for those ~300 species. To begin, we have a high-level overview of how PEP classifies the covered species:

<div class="narrow-tab">
```{r echo=FALSE}
PEP <- readRDS("data/HI_PEP_2017-03-16.rds")
extrp_wild <- filter(PEP, island_designation == "PEP-EXTIRPATED" |
                       island_designation == "PEP-EXTIRPATED?")
table(PEP$island_designation) %>%
  as_data_frame() %>%
  rename(PEP_class = Var1) %>%
  filter(PEP_class != "N/A") %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div> 

The numbers from PEP indicate the scale of the challenge Hawaiian plants face: nearly 300 species covered in this extinction prevention program, and dozens that are extinct or extirpated. Note the refinement of extinct in the wild vs. extinct from the planet, which we will return to later. But first, what are the PEP species that are thought extinct? ("NL" = not listed)

<div class="narrow-tab">
```{r pep_ext_table, echo = FALSE, message = FALSE}
pep_extinct <- filter(PEP, island_designation == "PEP-EXTINCT?")
list_stat <- c("NL", rep("E", 5), "NL", rep("E", 3), "NL", "E", rep("NL", 5),
               "E", NA, "E")
pep_extinct$status <- list_stat
pep_extinct$distribution <- str_replace_all(
  pep_extinct$distribution,
  "`", "'"
)
kable(pep_extinct) %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

The PEP tabulation places 20 Hawaiian plants they have worked with in the EXTINCT?`r tags$sup(tags$span(style="background-color:red;color:#fff;border-radius:2px;padding:1px 5px", "NOTE (click)")) %>% bs_embed_popover(title = "NOTE", content = "Appropriately, PEP uses '?' to denote that they are not declaring species extinct or that there is uncertainty.")` category. Of these, eleven are listed (endangered), and their listing rules or recently released five-year reviews suggest extinction. 

### Declared extinct

In addition to the species identified above, we know that some species have been delisted because they have been formally designated as extinct.

<div class="narrow-tab">
```{r delist_ext, echo = FALSE, message = FALSE}
delisted <- readr::read_tsv("data/delisted_species_2018-01-03.tsv")
names(delisted) <- c("date_delisted", "species", "date_listed", "reason_delisted",
                   "dom_for", "service")
dl_ext <- filter(delisted, reason_delisted == "Extinct")
dl_ext$species <- str_replace_all(dl_ext$species, "[ ]+- Wherever found", "")
select(dl_ext, c(2,3,1)) %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover")
```
</div>

```{r n_extinct_calcs, echo = FALSE}
pep_ext_list <- filter(pep_extinct, status == "E")
n_ext <- length(spp) + length(dl_ext$species) + length(ext_ach$species) +
         length(pep_ext_list$species)
```

### Extinct in the wild

We have focused on species that are completely extinct, but not those that are extinct in the wild. Even though we do not include those species in the tally, we note that these often number just a few individuals in propagation or captivity--very close to complete extinction. We identified `r length(ext_wild[[1]])` such species in our scans--_`r paste(ext_wild[[1]], collapse=", ")`_--but the number is likely higher. For example, the PEP data indicated 14 Hawaiian plants that are extinct in the wild.

<div class="main-result">
**Available data indicate `r n_ext` (`r 100*round(n_ext/dim(tne)[1], 3)`%) current or formerly ESA-listed species are thought to be extinct.** This is substantially higher than commonly recognized.
</div>

## Causes of extinction

In the course of compiling the list of likely extinct species, we read about the causes of these species' extinction. At this time we can't "conclusively" determine the causes of extinction of all `r n_ext` species identified above, but we think it is worth noting three basic categories of extinction causes we identified, from most to least frequent`r tags$sup(tags$span(style="background-color:red;color:#fff;border-radius:2px;padding:1px 5px", "NOTE (click)")) %>% bs_embed_popover(title = "NOTE", content = "We do not include counts of the number of species for each category because the available information is too spotty to ensure the counts are accurate; instead, we simply order by our best estimate of frequency. This list should be considered an initial observation worthy of further investigation rather than the final word on the causes of ESA extinctions.")`:

<div id="cause_list">
1. _Invasive species_ were the factor we noticed most commonly in discussions of extinction, with the frequency driven by the _Achatinella_ snails and Hawaiian plants. The culprits include feral hogs, predatory snails, avian malaria, and a multitude of plants. Regardless the mechanism by which the invasives affected the listed species--trampling, predation, disease, competition--the end result was the same: species driven to extinction, or so compromised that they became susceptible to factors such as natural "accidents."

2. _Natural accidents_ were a surprisingly common theme in the narratives of extinction. _Hibiscadelphus woodii_, for example, was only ever known from four plants, three of which were taken out by a rockslide and the fourth died some years later. (Efforts were made to propagate the species, but were unsuccessful.) Several plants are believed to have been lost to Hurricane Iniki in 1992, including _Dubautia kenwoodii_. These sorts of natural accidents (and other stochastic events) are the reason why small population size is considered a significant extinction risk, for example, under the [five listing factors of the ESA](http://www.nmfs.noaa.gov/pr/listing/). We are unable to determine the extent to which invasive species or "unnatural processes" increased the susceptibility to natural accidents, but those are likely situations.

3. _"Unnatural processes"_ were characteristic for some of the commonly recognized species on the list. The eastern cougar and the Eskimo Curlew almost certainly fell victim to overharvest. Others, like the Maryland darter and Bachman's Warbler, most likely fell victim to anthropogenic habitat destruction and degradation. Similarly, several mussels in the Southeast we likely driven to extinction by a combination of abiotic habitat modification such as silt and pollution and (perhaps) biotic changes such as loss of fish hosts. Even though these types of processes were likely responsible for some extinctions, what stood out was how they have been relatively infrequent.
</div>

The pattern of these causes raises some important questions about whether the ESA--or any law--could have prevented many of these likely extinctions. How can laws that typically focus on prohibiting harmful acts be used to address well-established and harmful invasive species and accidents? By extension, does messaging that ignores the different types of threats and the different mechanisms that might ameliorate the threats set unrealistic expectations for the ESA? 

<div class="main-result">
**These observations raise the question of whether any law can deal with some of the most common causes of (past) extinctions, and how we message law's capabilities.**
</div>

## Conclusion

Our analysis points to two basic conclusions about messaging the effectiveness and expectation of the ESA. First, conservationists should not continue using the "99% effective" phrase to discuss the ESA's success: the data indicate the percent extant is likely closer to 95%. Sometimes the statistic is rephrased as "only 1% of listed species have been declared extinct," which is technically true but deliberately misleading. (And one can even argue that it is "disrespectful" to those species that have likely gone extinct to not recognize the loss.) However, even with the better estimate or the technically correct phrasing, we believe that the ESA and conservation are better served by not focusing on the lower bar of percent of species extant. Instead, we believe that a focus on recovery and conservation well-above near-extinction.

Second, conservationists should discuss causes of extinction--and threats to extant listed species--in proportion to their frequency and importance rather than just focusing on the manageable threats. The latter may make for simple success stories, but distorts expectations of what the ESA can do. This is part-and-parcel with messaging that implies simple listing a species under the ESA is sufficient to ensure its conservation. Prior to working on this analysis, we would have guessed overharvesting and habitat destruction were the primary causes of extinction of ESA-listed species...and we would have been wrong.

### Coda

Conducting this analysis hasn't been the most uplifting exercise: not much good news in here. But as we cautioned from the beginning, declaring extinction is fundamentally hard and is a step taken only with great care. There is no better example to illustrate why we must take care--and why we do not think the 89 species identified here should be declared extinct until species experts deem it appropriate--than _Melicope quadrangularis_. This Hawaiian tree was discovered in 1909, then "lost" until 1991, when a population was found again. Then Hurricane Iniki struck in 1992 and for 20 years it was thought that species had been wiped out because that single population had been destroyed by the hurricane. When we first did this analysis in mid-2017, the most recent 5-year review (2010) indicated this history and we identified it as a likely extinct species in our list. But the [2017 5-year review](https://ecos.fws.gov/docs/five_year_review/doc5340.pdf), which we reviewed as part of finalizing this analysis, presented hopeful new information: a stand of four _M. quadrangularis_ was found in another location in 2013, followed by other populations in subsequent years. Efforts to find and propagate the species are continuing by PEP, which is great news. 


## Methods {#methods}

We used a two-step process to identify the likely extinct species. First, we used a coarse filter to search across all ~1,400 five-year reviews with seven [regular expression](https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf) patterns:

<div style="font-size:smaller">
1. `"(possibly|may be|have become) extinct"`
2. `"(certainly being|probably|is|likely|probably being) extinct"`
3. `"no (populations|individuals)( are)* known"`
4. `"no( known| living)* individuals"`
5. `"extinct in( the)* wild"`
6. `"last (seen|observed|found) in [0-9oOiI]+"`
7. `"functionally extinct"`
</div>
    
In these patterns, "|" means "or" for the set of words inside parentheses. For example, the first pattern would match "possibly extinct," "may be extinct," or "have become extinct." The "\*" means the preceding letter or word matches zero or more times, whereas "+" means the preceding word/letter(s) must match one or more times. We checked all seven patterns against all 1,385 five-year reviews we collected from [ECOS](https://ecos.fws.gov) and [NMFS's recovery site](http://www.nmfs.noaa.gov/pr/recovery/plans.htm).

A basic search returned `r dim(ext_pat)[1]` matches to [the set of candidate phrase patterns](#phrase_patterns), including patterns that indicate a higher likelihood of extinction such as: 

<div class="narrow-tab">
```{r patt_2, echo = FALSE} 
t4 <- filter(ext_pat, patt == 2) %>% 
  select(mats) %>% 
  data.frame() %>%
  head()
kable(paste("-", t4$mats))
```
</div>

Perusing these example pattern matches, we find - as expected - both true positives (i.e., the meaning of the matched sentence comports with the idea we have in mind) and false positives (e.g., local extinctions or species previously believed extinct but since re-discovered). We manually checked each 5-year review with an extinction phrase match to remove the false-positives. 

The _Achatinella_ snails are taxonomically complex and all of their five-year reviews have been combined in a [single document](https://ecos.fws.gov/docs/five_year_review/doc3903.pdf), so we compiled their results manually for all species. In addition, we checked the NatureServe Explorer and IUCN Red List web sites for any additional information for the snails. 

We downloaded the 2017-03-16 list from PEP's website then used Adobe's online extraction service to render the tables as a spreadsheet for processing. We checked the PEP species against the list of threatened and endangered species to remove non-listed species from consideration. We also checked available documents (e.g., the _Federal Register_) to get a better picture of why/how the species went extinct.

We downloaded the list of delisted species from FWS's [ECOS website](https://ecos.fws.gov/ecp0/reports/delisting-report) in January, 2018, and included only those species for which the "Reason Delisted" was Extinct.


