---
title: "Extinct and declining ESA-listed species"
author: "Defenders of Wildlife"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    css: custom.css
    df_print: paged
    fig_caption: yes
    fig_width: 7
    fig_height: 5
    highlight: tango
    toc: true
    toc_depth: 4
    toc_float: true
editor_options: 
  chunk_output_type: console
---

<script async defer src="https://hypothes.is/embed.js"></script>

```{r setup, echo = FALSE, message=FALSE}
library(dplyr)
library(ecosscraper)
library(fuzzyjoin)
library(ggplot2)
library(parallel)
library(rio)
library(stringr)
library(tokenizers)
library(us.geonames)

src <- "~/Downloads/five_year_review"
fils <- list.files(src, full.names = TRUE, recursive = TRUE)

### Five-year review tables from ECOS 
# load("/Users/jacobmalcom/Downloads/ECOS_species_tables_2016-12-17.rda")
# fiveyr_table$doc <- basename(fiveyr_table$Doc_Link) %>% 
#   str_replace_all(pattern = "pdf$", replacement = "txt")
# saveRDS(fiveyr_table, file = "data/fiveyr_table_2016-12-17.rds")
fiveyr_table <- readRDS("data/fiveyr_table_2016-12-17.rds")

### County occurrences
# cnt_occ <- readr::read_tsv("~/Downloads/EndSp_county_occurrences.tsv")
# saveRDS(cnt_occ, file = "data/ESA_county_occurrences.rds")
cnt_occ <- readRDS("data/ESA_county_occurrences.rds")

### Recovery report to Congress
# rec_rep <- readxl::read_excel(
#   "~/Downloads/Recovery_Report_FY2013-2014_tabs.xlsx"
# )
# rec_rep$date_listed <- as.Date(rec_rep$date_listed)
# 
# rec_rep$date_5yr <- ifelse(
#   rec_rep$date_5yr == "N/A", 
#   NA, 
#   as.numeric(rec_rep$date_5yr)
# ) %>% as.Date(origin = "1899-12-30")
# 
# rec_rep$plan_date <- ifelse(
#   rec_rep$plan_date == "N/A", 
#   NA, 
#   as.numeric(rec_rep$plan_date)
# ) %>% as.Date(origin = "1899-12-30")
# 
# rec_rep$fiveyr_rec <- str_replace_all(rec_rep$fiveyr_rec, "\\*$", "") %>%
#   str_replace_all("species", "Species")
# rec_rep$species <- str_replace_all(rec_rep$species, "\r\n", " ")
# rio::export(rec_rep, file = "~/Downloads/ESA_recovery_report_2013-2014.tsv")
# saveRDS(rec_rep, file = "data/ESA_recovery_report_2013-2014.rds")
rec_rep <- readRDS("data/ESA_recovery_report_2013-2014.rds")

# PEP <- readxl::read_excel("~/Downloads/HI_TE_plants_pep_list_20170316.xlsx")
# names(PEP) <- c("species", "distribution", "island_designation")
# PEP$island_designation <- ifelse(
#   PEP$island_designation == "PEP-Extinct in wild",
#   "PEP-Extinct in Wild",
#   PEP$island_designation
# )
# saveRDS(PEP, file = "data/HI_PEP_2017-03-16.rds")
PEP <- readRDS("data/HI_PEP_2017-03-16.rds")

### Recovery plan BAS from five-year review
# txt <- paste(readLines("~/Downloads/fiveyr_recplan_BAS.txt"), collapse = " ")
# parts <- str_split(txt, pattern = "--")[[1]]
# fnames <- str_match(parts, pattern = "^[ ]*[a-zA-Z0-9-_]*\\.txt") %>%
#   str_replace(pattern = "^ ", replacement = "")
# BAS_df <- data_frame(file = fnames, text = parts)
# BAS_df$doc <- basename(BAS_df$file)
# BAS_df$yes_checks <- grepl(parts, pattern = "X[-_ ]*Yes")
# BAS_df$no_checks <- grepl(parts, pattern = "X[-_ ]*No")
# BAS_df$anti_check <- !grepl(parts, pattern = "X[-_ ]*")
# BAS_df$ac_yes <- ifelse(
#   BAS_df$anti_check == TRUE & grepl(BAS_df$text, pattern = "Yes[.]*"),
#   TRUE,
#   FALSE
# )
# BAS_df$ac_no <- ifelse(
#   BAS_df$anti_check == TRUE & grepl(BAS_df$text, pattern = "No[.]*"),
#   TRUE,
#   FALSE
# )
# BAS_df$recplan_BAS <- ifelse(
#   BAS_df$yes_checks | BAS_df$ac_yes, 
#   TRUE, 
#   FALSE
# )
# saveRDS(BAS_df, file = "data/recplan_BAS_fiveyr.rds")
BAS_df <- readRDS("data/recplan_BAS_fiveyr.rds")

### Extinct species from 5-year reviews
ext_5yr <- readxl::read_excel("~/Downloads/extinct_patterns_raw.xlsx")
saveRDS(ext_5yr, file = "data/extinct_manual_check_2017-09-05.rds")
ext_5yr <- readRDS("data/extinct_manual_check_2017-09-05.rds")
```

```{r proc, echo = FALSE}
terms <- c(
  "(possibly|may be|have become) extinct",
  "(certainly being|probably|is|likely|probably being) extinct",
  "no (populations|individuals)( are)* known",
  "no( known| living)* individuals",
  "extinct in( the)* wild",
  "last (seen|observed|found) in [0-9oOiI]+",
  "functionally extinct"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- get_text(f)
  sent <- tokenize_sentences(text, lowercase = TRUE, simplify = TRUE)
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(terms)) {
    mat <- grep(sent, pattern = terms[i], fixed = FALSE, value = TRUE)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

t1 <- parallel::mclapply(fils, proc_file, mc.cores = 3, mc.preschedule = FALSE)
ext_pat <- bind_rows(t1) %>% filter(!is.na(mats))
last_seen <- filter(ext_pat, patt == 6)
func_exct <- filter(ext_pat, patt == 7)
```

The number of 5-year reviews that include an extinction phrase, which is a 
lower bound for the number of species whose review includes an extinction
phrase:

```{r echo=FALSE}
length(unique(ext_pat$doc))
```

I would like to know which species this includes and where those species are
found, so will do a couple of joins.

```{r echo = FALSE}
w_spp <- left_join(ext_pat, fiveyr_table, by = "doc")
w_co <- left_join(w_spp, cnt_occ, by = c("Species" = "SCI"))

spp_w_st <- distinct(w_co, Species, STATE)
VA_spp <- filter(w_co, STATE == "Virginia")
unique(VA_spp$Species)
```

# Recovery plans: best available science?

```{r recplan_BAS}
BAS_miss <- setdiff(fiveyr_table$doc, BAS_df$doc)
```

# Listing status recommendations

```{r echo=FALSE}
patts <- c(
  "[xX][-_ ]*( Yes, )*[dD]ownlist",
  "[xX][-_ ]*( Yes, )*[rR]eclassify from Endangered to Threatened",
  "[xX][-_ ]*( Yes, )*[uU]plist",
  "[xX][-_ ]*( Yes, )*[rR]eclassify from Threatened to Endangered",
  "[xX][-_ ]*( Yes, )*[dD]elist",
  "[xX][a-z -_]*( No, )*[nN]o [cC]hange",
  "Recommended [cC]lassification[[:print:]]+: Threatened",
  "Recommended [cC]lassification[[:print:]]+: Endangered",
  "Recommended [cC]lassification[[:print:]]+: No change",
  "[xX][-_ ]*Extinct",
  "[xX][-_ ]*Recovery",
  "[xX][-_ ]*Original data"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- readLines(f) %>% str_replace_all("[ ]{2,}", " ")
  print(nchar(paste(text, collapse = "")))
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(patts)) {
    mat <- str_extract_all(text, pattern = patts[i], simplify = TRUE)
    mat <- mat[mat != ""]
    if(length(mat) > 0) print(mat)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

t1 <- mclapply(fils, proc_file, mc.cores = 3, mc.preschedule = FALSE)
stat_rec <- bind_rows(t1) %>% 
  filter(!is.na(file)) %>% 
  distinct(doc, mats, .keep_all = TRUE)

stat_rec$rec <- ifelse(
  stat_rec$patt == 1 & is.na(stat_rec$mats),
  NA,
  ifelse(
    stat_rec$patt == 1 | stat_rec$patt == 2,
    "downlist",
    ifelse(
      stat_rec$patt == 3 | stat_rec$patt == 4,
      "uplist",
      ifelse(
        stat_rec$patt == 5,
        "delist",
        ifelse(
          stat_rec$patt == 6 | stat_rec$patt == 9,
          "no change",
          ifelse(
            stat_rec$patt == 7,
            "Unk change, T",
            "Unk change, E"
          )
        )
      )
    )
  )
)
stat_2 <- distinct(stat_rec, doc, rec, .keep_all = TRUE) %>%
  arrange(-patt, mats)
stat_2$dups <- duplicated(stat_2$doc)
stat_3 <- filter(stat_2, dups == FALSE) %>%
  arrange(doc)
stat_3$delist_bc <- ifelse(
  stat_3$patt == 10,
  "extinct",
  ifelse(
    stat_3$patt == 11,
    "recovery",
    ifelse(
      stat_3$patt == 12,
      "error",
      NA
    )
  )
)
stat_rec <- left_join(stat_3, fiveyr_table, by = "doc") %>%
  select(-dups) %>%
  distinct(doc, .keep_all = TRUE)

delist <- filter(stat_rec, rec == "delist")
uplist <- filter(stat_rec, rec == "uplist")
downlist <- filter(stat_rec, rec == "downlist")

table(stat_rec$rec, useNA = "always")
table(stat_rec$delist_bc)
```

## Compare to biennial report

```{r}
table(rec_rep$fiveyr_rec)
spp_df <- data_frame(species = unique(cnt_occ$SCI))
domest <- get_TECP_table() %>% filter_domestic() %>% filter_listed()
rr2 <- fuzzy_left_join(
  rec_rep,
  domest,
  by = c("species" = "Scientific_Name"),
  match_fun = function(v1, v2) {
    stringr::str_detect(string = v1, pattern = v2)
  }
)

no_sci <- filter(rr2, is.na(Scientific_Name))

delist_ext <- filter(rec_rep, fiveyr_rec == "Delist due to extinction")
```

# PEP

The [Plant Extinction Prevention (PEP) Program](http://www.pepphi.org/) in 
Hawaii tracks the status of imperiled plants more closely than FWS documents,
and should probably be considered authoritative for species covered by PEP.

```{r}
as_data_frame(table(PEP$island_designation))
```

How many HI plant species have been extirpated from one or more islands?

```{r}
extrp_wild <- filter(PEP, island_designation == "PEP-EXTIRPATED" |
                       island_designation == "PEP-EXTIRPATED?")
length(unique(extrp_wild$species))
```

And the distribution among species and islands?

```{r}
spXis_tab <- table(extrp_wild$species, extrp_wild$distribution)
colSums(spXis_tab)
sort(rowSums(spXis_tab), decreasing = TRUE)
```

Now let's join the PEP table with counties data to get an idea of how many
species are not covered by PEP:

```{r}
hi_cnt <- filter(cnt_occ, STATE == "Hawaii")
pep_miss <- setdiff(unique(hi_cnt$SCI), PEP$species)
length(pep_miss)
```

So more are missing than are covered. How many have 5y reviews and what is the 
distribution of those review dates?

```{r}
pep_miss_5y <- filter(fiveyr_table, fiveyr_table$Species %in% pep_miss)
pep_miss_5y$Date <- as.Date(pep_miss_5y$Date)
p <- qplot(
  data = pep_miss_5y, 
  x = pep_miss_5y$Date, 
  geom = "histogram", 
  bins = 9
)
plotly::ggplotly(p)
```

# Post manual checks

```{r }
dist <- filter(ext_5yr, true_positive == "T") %>% 
  distinct(Species, .keep_all = TRUE) %>%
  filter(!is.na(Date))

comments <- ext_5yr %>%
  distinct(Species, comment, .keep_all = TRUE)
comments$comment %>%
  table() %>% 
  sort(decreasing = TRUE)

# geography of extinction
ext_geo <- left_join(dist, cnt_occ, by = c("Species" = "SCI"))
ext_geo_state <- distinct(ext_geo, Species, STATE, .keep_all = TRUE)
ext_geo$co_st <- paste0(ext_geo$NAME, ", ", ext_geo$STATE)
ext_geo_st_co <- distinct(ext_geo, Species, co_st, .keep_all = TRUE)
table(ext_geo_state$STATE) %>% sort(decreasing = TRUE)
table(ext_geo_st_co$co_st) %>% sort(decreasing = TRUE) %>% head(15)

# add map...

```

# recovery potential

```{r echo=FALSE}
patts <- c(
  "potential for recovery",
  "recovery potential"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- readLines(f) %>% str_replace_all("[ ]{2,}", " ") %>% tolower()
  print(nchar(paste(text, collapse = "")))
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(patts)) {
    mat <- grep(x = text, pattern = patts[i], value = TRUE)
    mat <- mat[mat != ""]
    if(length(mat) > 0) print(mat)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

potent <- mclapply(fils, proc_file, mc.cores = 3, mc.preschedule = FALSE)
stat_potent <- bind_rows(potent) %>% 
  filter(!is.na(file)) %>% 
  filter(!is.na(mats)) %>% 
  distinct(doc, mats, .keep_all = TRUE)

low_potent <- filter(
  stat_potent, 
  grepl(stat_potent$mats, 
        pattern = "low recovery potential|low potential (of|for) recovery")) %>%
  left_join(low_potent, fiveyr_table, by = "doc")
length(unique(hi_potent$Species))

hi_potent <- filter(
  stat_potent, 
  grepl(
    stat_potent$mats, 
    pattern = "high recovery potential|high potential (of|for) recovery")
  ) %>%
  left_join(hi_potent, fiveyr_table, by = "doc")
length(unique(hi_potent$Species))

rr2$RPN_num <- str_replace_all(rr2$RPN, "C$", "")
rr2$RPN_conflict <- str_detect(rr2$RPN, "C$")
table(rr2$RPN_num)

low_rec_pot_BR <- filter(rr2, RPN_num %in% c(4,5,6,10,11,12,16,17,18))
hi_rec_pot_BR <- filter(rr2, !(RPN_num %in% c(4,5,6,10,11,12,16,17,18)))
rr2$recov_potential <- ifelse(
  rr2$RPN_num %in% c(4,5,6,10,11,12,16,17,18),
  "low", "high"
)
rr2$threat_deg <- ifelse(
  rr2$RPN_num %in% c(1:6),
  "high",
  ifelse(rr2$RPN_num %in% c(7:12),
         "medium",
         "low")
)
plotly::ggplotly(qplot(data = rr2, x = recov_potential, geom = "bar") +
  facet_grid(. ~ threat_deg))

# setdiff(low_rec_pot_BR$Scientific_Name, low_potent$Species)
intersect(low_rec_pot_BR$Scientific_Name, hi_potent$Species) %>%
  setdiff(low_potent$Species)
  
intersect(low_rec_pot_BR$Scientific_Name, low_potent$Species) 

## How many 'extinct' species are high recovery likelihood...
RTC_ex <- filter(rr2, Scientific_Name %in% dist$Species)
qplot(x = as.factor(as.numeric(RTC_ex$RPN_num))) +
  labs(x = "RPN") +
  ggthemes::theme_hc()
RTC_high_rec <- filter(RTC_ex, RPN_num == 2 | RPN_num == 3)
```

# Pop declines

```{r}
patts <- c(
  "population declin",
  "population increas"
)

proc_file <- function(f) {
  cat(paste("\nProcessing:", basename(f), "\n"))
  text <- readLines(f) %>% str_replace_all("[ ]{2,}", " ") %>% tolower()
  print(nchar(paste(text, collapse = "")))
  res <- data_frame(file = NA, patt = NA, doc = NA, mats = NA)
  for(i in 1:length(patts)) {
    mat <- grep(x = text, pattern = patts[i], value = TRUE)
    mat <- mat[mat != ""]
    if(length(mat) > 0) print(mat)
    nhit <- length(mat)
    if(nhit > 0) {
      resi <- data_frame(file = rep(f, nhit), 
                         doc = rep(basename(f), nhit),
                         patt = rep(i, nhit), 
                         mats = mat)
    } else {
      resi <- data_frame(file = f, patt = i, doc = basename(f), mats = NA)
    }
    res <- rbind(res, resi)
  }
  return(res)
}

pop_trend <- mclapply(fils[1:50], proc_file, mc.cores = 3, mc.preschedule = FALSE)
pop_trend_df <- bind_rows(pop_trend) %>% 
  filter(!is.na(file)) %>% 
  filter(!is.na(mats)) %>% 
  distinct(doc, mats, .keep_all = TRUE)


```



